---
title: "Natural Selection Signatures in Human Populations: A Bayesian Allele Frequency Analysis"
subtitle: "Preliminary Analysis — 1000 Genomes Project Phase 3"
author: "Cole M."
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 9, fig.height = 5)
```

---

## 1. Research Question & Background

**What are we doing?**
We use a Bayesian Beta-Binomial model to estimate allele frequencies for 12 variants known to be under natural selection, then ask: do these loci show statistically greater population differentiation than a background of common, neutrally-evolving variants? We also examine the haplotype structure around each locus as an independent line of evidence for recent selective sweeps.

**Data source:** [1000 Genomes Project Phase 3](https://www.internationalgenome.org/) provides allele counts from 2,504 individuals organized into five superpopulations (AFR, AMR, EAS, EUR, SAS). Allele counts and phenotype annotations are retrieved in real time via the [Ensembl REST API](https://rest.ensembl.org/).

> **Note on phenotype data:** The 1000 Genomes Project does **not** include individual-level phenotype measurements — it is a population-genetics resource. Phenotype/trait associations shown in Section 4 come from linked databases (GWAS Catalog, ClinVar, OMIM) accessed through Ensembl.

**The 12 focal alleles** span four superpopulations and cover both positive and balancing selection:

| SNP | Gene | Trait | Selection type | Primary pop |
|-----|------|-------|----------------|-------------|
| rs4988235 | *LCT* | Lactase persistence (milk digestion) | Positive | EUR |
| rs2814778 | *DARC/ACKR1* | Duffy-null, malaria resistance | Positive | AFR |
| rs3827760 | *EDAR* | Hair thickness / sweat gland morphology | Positive | EAS |
| rs1426654 | *SLC24A5* | Reduced skin pigmentation | Positive | EUR/SAS |
| rs334 | *HBB* | Sickle cell / malaria resistance | **Balancing** | AFR |
| rs17822931 | *ABCC11* | Dry earwax / reduced body odor | Positive | EAS |
| rs16891982 | *SLC45A2* | Reduced skin pigmentation | Positive | EUR |
| rs12913832 | *HERC2/OCA2* | Blue eye color | Positive | EUR |
| rs1805007 | *MC1R* | Red hair / reduced pigmentation | Positive | EUR |
| rs1229984 | *ADH1B* | Alcohol metabolism (fast oxidizer allele) | Positive | EAS |
| rs1800414 | *OCA2* | East Asian skin depigmentation | Positive | EAS |
| rs12203592 | *IRF4* | Freckling / hair & skin pigmentation | Positive | EUR |

**Background panel (10 SNPs):** Common GWAS variants without documented strong directional selection (MTHFR, COMT, BDNF, DRD2, PPARG, APOE×2, TP53, FTO, GSTP1) serve as a neutral comparison.

---

## 2. Setup

```{r packages}
required_pkgs <- c("httr", "jsonlite", "dplyr", "tidyr",
                   "ggplot2", "patchwork", "knitr", "scales")
new_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[,"Package"])]
if (length(new_pkgs)) install.packages(new_pkgs, repos = "https://cloud.r-project.org")

library(httr); library(jsonlite); library(dplyr); library(tidyr)
library(ggplot2); library(patchwork); library(knitr); library(scales)
```

```{r constants}
# 1000 Genomes Phase 3 haplotype counts (2 × individuals)
N_HAP <- c(AFR = 1322, AMR = 694, EAS = 1008, EUR = 1006, SAS = 978)

POP_COLORS <- c(AFR = "#E41A1C", AMR = "#FF7F00",
                EAS = "#4DAF4A", EUR = "#377EB8", SAS = "#984EA3")

# One representative subpopulation per superpop for LD queries
LD_POP_CODES <- c(
  AFR = "1000GENOMES:phase_3:YRI",
  AMR = "1000GENOMES:phase_3:MXL",
  EAS = "1000GENOMES:phase_3:CHB",
  EUR = "1000GENOMES:phase_3:CEU",
  SAS = "1000GENOMES:phase_3:GIH"
)

# Focal (selected) allele for each validation SNP
VALIDATION_SNPS <- data.frame(
  rsid         = c("rs4988235","rs2814778","rs3827760","rs1426654",
                   "rs334",    "rs17822931","rs16891982","rs12913832",
                   "rs1805007","rs1229984", "rs1800414", "rs12203592"),
  gene         = c("LCT",  "DARC",  "EDAR",   "SLC24A5",
                   "HBB",  "ABCC11","SLC45A2","HERC2",
                   "MC1R", "ADH1B", "OCA2",   "IRF4"),
  focal_allele = c("A","T","A","A",
                   "T","A","C","A",
                   "T","G","A","C"),
  sel_type     = c("positive","positive","positive","positive",
                   "balancing","positive","positive","positive",
                   "positive","positive","positive","positive"),
  selection_pop = c("EUR","AFR","EAS","EUR/SAS",
                    "AFR","EAS","EUR","EUR",
                    "EUR","EAS","EAS","EUR"),
  ld_query_pop  = c("EUR","AFR","EAS","EUR",
                    "AFR","EAS","EUR","EUR",
                    "EUR","EAS","EAS","EUR"),
  stringsAsFactors = FALSE
)

# Background panel — 10 common SNPs without strong directional selection
BACKGROUND_RSIDS <- c(
  "rs1801133", "rs4680",   "rs6265",   "rs1800497",
  "rs1801282", "rs7412",   "rs429358", "rs1042522",
  "rs9939609", "rs1695"
)
```

---

## 3. Data Retrieval

### 3.1 API helper functions

```{r api_functions}
# Fetch allele + phenotype data for a single variant from Ensembl REST API
fetch_variant <- function(rsid, pause_sec = 0.35) {
  url <- paste0("https://rest.ensembl.org/variation/human/", rsid,
                "?pops=1;phenotypes=1")
  Sys.sleep(pause_sec)
  resp <- tryCatch(
    GET(url, add_headers(`Content-Type` = "application/json"), timeout(30)),
    error = function(e) NULL
  )
  if (is.null(resp) || status_code(resp) != 200) return(NULL)
  content(resp, as = "parsed", type = "application/json")
}

# Extract allele counts per superpopulation
parse_variant <- function(api_data, rsid, focal_allele = NULL) {
  if (is.null(api_data) || is.null(api_data$populations)) return(NULL)
  target <- if (!is.null(focal_allele)) focal_allele else api_data$minor_allele
  if (is.null(target)) return(NULL)

  pops_df  <- bind_rows(lapply(api_data$populations, as.data.frame))
  superpops <- c("AFR","AMR","EAS","EUR","SAS")

  bind_rows(lapply(superpops, function(sp) {
    entries   <- pops_df[pops_df$population == paste0("1000GENOMES:phase_3:", sp), ]
    if (nrow(entries) == 0) return(NULL)
    n         <- sum(entries$allele_count, na.rm = TRUE)
    focal_row <- entries[entries$allele == target, ]
    alt_count <- if (nrow(focal_row) > 0) focal_row$allele_count[1] else 0L
    data.frame(rsid = rsid, population = sp,
               alt_count = as.integer(alt_count), n = as.integer(n),
               freq_obs  = if (n > 0) alt_count / n else NA_real_,
               stringsAsFactors = FALSE)
  }))
}

# Extract phenotype annotations from the API response
parse_phenotypes <- function(api_data, rsid, gene_label) {
  if (is.null(api_data) || is.null(api_data$phenotypes)) return(NULL)
  phenos <- bind_rows(lapply(api_data$phenotypes, function(p) {
    data.frame(
      rsid        = rsid,
      gene        = gene_label,
      trait       = if (!is.null(p$trait))       p$trait       else NA_character_,
      source      = if (!is.null(p$source))      p$source      else NA_character_,
      risk_allele = if (!is.null(p$risk_allele)) p$risk_allele else NA_character_,
      stringsAsFactors = FALSE
    )
  }))
  phenos[!is.na(phenos$trait) & phenos$trait != "", ]
}
```

### 3.2 Fetch all SNPs

```{r fetch_data, cache=TRUE}
val_data_list <- list()
pheno_list    <- list()

for (i in seq_len(nrow(VALIDATION_SNPS))) {
  rs  <- VALIDATION_SNPS$rsid[i]
  fa  <- VALIDATION_SNPS$focal_allele[i]
  api <- fetch_variant(rs)

  parsed <- parse_variant(api, rs, fa)
  if (!is.null(parsed)) {
    parsed$gene  <- VALIDATION_SNPS$gene[i]
    parsed$panel <- "validation"
    val_data_list[[i]] <- parsed
  }
  pheno <- parse_phenotypes(api, rs, VALIDATION_SNPS$gene[i])
  if (!is.null(pheno)) pheno_list[[i]] <- pheno
}
val_data   <- bind_rows(val_data_list)
pheno_data <- bind_rows(pheno_list)

bg_data_list <- list()
for (i in seq_along(BACKGROUND_RSIDS)) {
  rs  <- BACKGROUND_RSIDS[i]
  api <- fetch_variant(rs)
  parsed <- parse_variant(api, rs, focal_allele = NULL)
  if (!is.null(parsed)) {
    parsed$gene  <- rs
    parsed$panel <- "background"
    bg_data_list[[i]] <- parsed
  }
}
bg_data  <- bind_rows(bg_data_list)
all_data <- bind_rows(val_data, bg_data)

cat("Retrieved — validation SNPs:", n_distinct(val_data$rsid),
    "| background SNPs:", n_distinct(bg_data$rsid),
    "| phenotype records:", nrow(pheno_data), "\n")
```

---

## 4. Phenotype Associations

The 1000 Genomes Project provides **allele frequency data only** — no individual phenotype measurements exist in the dataset. The table below shows trait annotations retrieved from GWAS Catalog, ClinVar, and OMIM via the Ensembl API. These confirm that each validation SNP has documented functional consequences.

```{r phenotype_table}
if (nrow(pheno_data) > 0) {
  pheno_data %>%
    filter(!is.na(trait)) %>%
    distinct(gene, rsid, trait, source, risk_allele) %>%
    arrange(gene, source) %>%
    kable(col.names = c("Gene","rsID","Phenotype / trait","Database","Effect allele"),
          caption = "Phenotype associations for validation SNPs (from Ensembl-linked databases)")
} else {
  cat("No phenotype records returned from API in this run.\n")
}
```

```{r phenotype_count_plot, fig.height=4}
if (nrow(pheno_data) > 0) {
  pheno_data %>%
    filter(!is.na(trait)) %>%
    distinct(gene, rsid, trait, source) %>%
    count(gene, source) %>%
    ggplot(aes(x = reorder(gene, -n), y = n, fill = source)) +
    geom_col(color = "white", linewidth = 0.3) +
    labs(title = "Documented phenotype associations per validation SNP",
         subtitle = "Source: GWAS Catalog, ClinVar, and OMIM via Ensembl REST API",
         x = "Gene", y = "Number of trait associations", fill = "Database") +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

---

## 5. Bayesian Posterior Estimation

### 5.1 Model

Each population's allele count $x_i$ is modelled as Binomial, and we place a Beta(1,1) (uniform) prior on frequency $p_i$:

$$x_i \sim \text{Binomial}(n_i,\, p_i), \qquad p_i \sim \text{Beta}(1, 1)$$

The closed-form posterior is:

$$p_i \mid x_i \sim \text{Beta}(1 + x_i,\; 1 + n_i - x_i)$$

At 1000 Genomes sample sizes ($n > 600$ haplotypes per superpopulation), the posterior mean closely tracks the observed frequency. The Bayesian approach gives us a full probability distribution over $p$ — not just a point estimate — which we then use to quantify differentiation between populations.

### 5.2 Observed allele frequencies

```{r freq_table}
compute_posterior <- function(data, alpha_prior = 1, beta_prior = 1) {
  data %>%
    mutate(
      alpha_post = alpha_prior + alt_count,
      beta_post  = beta_prior  + (n - alt_count),
      post_mean  = alpha_post / (alpha_post + beta_post),
      post_sd    = sqrt((alpha_post * beta_post) /
                        ((alpha_post + beta_post)^2 * (alpha_post + beta_post + 1))),
      ci_lo      = qbeta(0.025, alpha_post, beta_post),
      ci_hi      = qbeta(0.975, alpha_post, beta_post)
    )
}

posterior_data <- compute_posterior(all_data)

# Wide-format observed frequency table
val_data %>%
  select(rsid, gene, population, freq_obs) %>%
  pivot_wider(names_from = population, values_from = freq_obs) %>%
  left_join(VALIDATION_SNPS %>% select(rsid, sel_type, selection_pop), by = "rsid") %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable(col.names = c("rsID","Gene","AFR","AMR","EAS","EUR","SAS",
                      "Selection type","Primary pop"),
        caption = "Observed focal-allele frequencies across 1000 Genomes superpopulations")
```

### 5.3 Posterior distributions

Each curve shows the full Beta posterior for one superpopulation. Wide separation between curves = strong population differentiation in allele frequency.

```{r posterior_plots, fig.height=14}
val_post <- posterior_data %>% filter(panel == "validation")
x_seq    <- seq(0.001, 0.999, length.out = 400)

plot_posterior_snp <- function(rsid_sel, gene_name) {
  snp_data <- val_post %>% filter(rsid == rsid_sel)
  dens_df  <- snp_data %>%
    rowwise() %>%
    mutate(dens = list(data.frame(p = x_seq,
                                  density = dbeta(x_seq, alpha_post, beta_post)))) %>%
    unnest(dens)

  ggplot(dens_df, aes(x = p, y = density, color = population, fill = population)) +
    geom_line(linewidth = 0.9) +
    geom_area(alpha = 0.12, position = "identity") +
    scale_color_manual(values = POP_COLORS) +
    scale_fill_manual(values  = POP_COLORS) +
    labs(title = paste0(gene_name, "  (", rsid_sel, ")"),
         x = "Allele frequency p", y = "Posterior density",
         color = "Population", fill = "Population") +
    theme_classic(base_size = 9) +
    theme(legend.position = "right", plot.title = element_text(size = 9, face = "bold"))
}

plots <- mapply(plot_posterior_snp,
                VALIDATION_SNPS$rsid, VALIDATION_SNPS$gene,
                SIMPLIFY = FALSE)

wrap_plots(plots, ncol = 3) +
  plot_annotation(
    title    = "Posterior Beta distributions — all 12 validation SNPs",
    subtitle = "Prior: Beta(1,1). Each curve = one 1000G superpopulation.",
    theme    = theme(plot.title = element_text(face = "bold", size = 12))
  )
```

**Key patterns:**

- **LCT (EUR):** Near-sweep in Europeans (~77%) vs. near-zero elsewhere — classic positive selection signature.
- **DARC (AFR):** T allele near-fixed in Africans (~97%) conferring malaria resistance via Duffy-null phenotype.
- **EDAR / ABCC11 (EAS):** Both near-fixed in East Asians, consistent with a shared sweep on chromosome 2.
- **HBB (balancing selection):** Sickle-cell allele maintained at intermediate frequency in AFR (~10–13%) — heterozygote advantage against malaria. No single population approaches fixation.
- **ADH1B / OCA2 (new EAS loci):** G allele of ADH1B and A allele of OCA2 should be elevated in East Asians relative to other populations.
- **IRF4 (new EUR locus):** C allele associates with freckling/pigmentation variation; expected EUR enrichment.

---

## 6. Are These Alleles Under Selection?

We measure **population differentiation** using the Bhattacharyya distance (BD) between pairs of posterior Beta distributions. BD = 0 means identical posteriors (no differentiation); higher BD means greater differentiation. We compare validation SNPs against the background panel.

```{r differentiation}
bhattacharyya_coef <- function(a1, b1, a2, b2) {
  if (any(is.na(c(a1,b1,a2,b2)))) return(NA_real_)
  tryCatch(
    integrate(function(x) sqrt(dbeta(x,a1,b1) * dbeta(x,a2,b2)),
              0, 1, subdivisions = 200L, rel.tol = 1e-6)$value,
    error = function(e) NA_real_)
}

pairwise_bd <- function(snp_post_data) {
  pops  <- snp_post_data$population
  pairs <- combn(length(pops), 2, simplify = FALSE)
  bind_rows(lapply(pairs, function(idx) {
    i <- idx[1]; j <- idx[2]
    bc <- bhattacharyya_coef(snp_post_data$alpha_post[i], snp_post_data$beta_post[i],
                              snp_post_data$alpha_post[j], snp_post_data$beta_post[j])
    data.frame(pop1 = pops[i], pop2 = pops[j],
               BD   = if (!is.na(bc) && bc > 0) -log(bc) else NA_real_)
  }))
}
```

```{r compute_bd, cache=TRUE}
snp_scores <- posterior_data %>%
  group_by(rsid, gene, panel) %>%
  group_split() %>%
  lapply(function(df) {
    pw      <- pairwise_bd(df)
    pw$rsid <- df$rsid[1]
    pw$gene <- df$gene[1]
    pw$panel <- df$panel[1]
    pw
  }) %>%
  bind_rows() %>%
  group_by(rsid, gene, panel) %>%
  summarise(max_BD  = max(BD,  na.rm = TRUE),
            mean_BD = mean(BD, na.rm = TRUE),
            .groups = "drop") %>%
  arrange(desc(max_BD))
```

### 6.1 SNP differentiation ranking

```{r rank_plot, fig.height=8}
rank_df <- snp_scores %>%
  mutate(label = ifelse(panel == "validation", gene, rsid)) %>%
  arrange(max_BD) %>%
  mutate(label = factor(label, levels = label))

ggplot(rank_df, aes(x = max_BD, y = label, color = panel, shape = panel)) +
  geom_segment(aes(x = 0, xend = max_BD, y = label, yend = label),
               color = "grey75", linewidth = 0.5) +
  geom_point(size = 3.5) +
  scale_color_manual(values = c(validation = "#D7191C", background = "#2C7BB6"),
                     labels = c("Validation (selection)", "Background (neutral)"),
                     name = NULL) +
  scale_shape_manual(values = c(validation = 17, background = 16), name = NULL) +
  labs(title = "Population differentiation: max pairwise Bhattacharyya distance",
       subtitle = "Validation SNPs (red) should cluster above background (blue) if they are under selection",
       x = "Max pairwise Bhattacharyya distance (BD)", y = NULL) +
  theme_classic(base_size = 11) +
  theme(axis.text.y = element_text(size = 9), legend.position = "bottom")
```

### 6.2 Validation vs. background: statistical test

```{r selection_test, fig.height=5}
val_scores <- snp_scores %>% filter(panel == "validation") %>% pull(max_BD)
bg_scores  <- snp_scores %>% filter(panel == "background")  %>% pull(max_BD)

# Standardize relative to background
bg_mean <- mean(bg_scores, na.rm = TRUE)
bg_sd   <- sd(bg_scores,   na.rm = TRUE)

snp_scores_z <- snp_scores %>%
  mutate(z_score    = (max_BD - bg_mean) / bg_sd,
         is_outlier = z_score > 2,
         label      = ifelse(panel == "validation", gene, rsid))

# Wilcoxon test: are validation SNPs more differentiated than background?
wt <- wilcox.test(val_scores, bg_scores, alternative = "greater")
cat(sprintf("Wilcoxon rank-sum (validation > background): W = %.0f, p = %.4f\n",
            wt$statistic, wt$p.value))

# Plot
ggplot(snp_scores, aes(x = panel, y = max_BD, color = panel)) +
  geom_boxplot(outlier.shape = NA, width = 0.35, linewidth = 0.8, fill = NA) +
  geom_jitter(aes(shape = panel), width = 0.12, size = 3.5, alpha = 0.85) +
  geom_text(data = snp_scores %>% filter(panel == "validation"),
            aes(label = gene), hjust = -0.25, size = 3, color = "black") +
  scale_color_manual(values = c(validation = "#D7191C", background = "#2C7BB6")) +
  scale_shape_manual(values = c(validation = 17, background = 16)) +
  scale_x_discrete(labels = c(validation = "Selection loci\n(n = 12)",
                               background = "Background\n(n = 10)")) +
  labs(title = "Max pairwise BD: selection loci vs. neutral background",
       subtitle = sprintf("Wilcoxon one-sided p = %.4f", wt$p.value),
       x = NULL, y = "Max pairwise Bhattacharyya distance") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none")
```

```{r z_score_table}
snp_scores_z %>%
  filter(panel == "validation") %>%
  select(label, max_BD, z_score, is_outlier) %>%
  arrange(desc(z_score)) %>%
  mutate(across(c(max_BD, z_score), ~ round(.x, 3))) %>%
  kable(col.names = c("Gene","Max BD","Z-score (vs. background)","Outlier (Z > 2)"),
        caption = "BD Z-scores for validation SNPs relative to the background distribution")
```

### 6.3 Published selection scores (reference)

Independent evidence from haplotype-based statistics confirms strong selection at the original four well-studied loci:

```{r published_scores}
data.frame(
  Gene      = c("LCT","DARC/ACKR1","EDAR","SLC24A5"),
  rsid      = c("rs4988235","rs2814778","rs3827760","rs1426654"),
  iHS_pop   = c("EUR (CEU)","AFR (YRI)","EAS (CHB)","EUR/SAS"),
  iHS_score = c("~5.8","~4.1","~6.2","~5.5"),
  XP_EHH    = c("EUR vs AFR ~6.0","AFR vs EUR ~5.8","EAS vs EUR ~7.1","EUR vs AFR ~5.5"),
  Reference = c("Voight et al. 2006","Hamblin & Di Rienzo 2000",
                "Fujimoto et al. 2008","Lamason et al. 2005")
) %>%
  kable(col.names = c("Gene","rsID","iHS pop","iHS score","XP-EHH","Reference"),
        caption = "Published haplotype-based selection scores (iHS > |2| = selection candidate; > |4| = strong outlier)")
```

---

## 7. Haplotype Structure

A key signature of **recent positive selection** is extended haplotype homozygosity (EHH): a newly swept allele has had insufficient time for recombination to break down the long haplotype on which it arose. This appears as high r² LD extending over hundreds of kilobases around the focal SNP. We query the Ensembl LD API within a 300 kb window around each validation SNP in its primary selection population.

```{r ld_functions}
fetch_ld <- function(rsid, population, window_kb = 300, r2_min = 0.1) {
  pop_full <- LD_POP_CODES[population]
  url <- paste0("https://rest.ensembl.org/ld/human/", rsid, "/",
                URLencode(pop_full, reserved = TRUE),
                "?window_size=", window_kb, "&r2=", r2_min)
  Sys.sleep(0.5)
  resp <- tryCatch(
    GET(url, add_headers(`Content-Type` = "application/json"), timeout(60)),
    error = function(e) NULL
  )
  if (is.null(resp) || status_code(resp) != 200) return(NULL)
  result <- content(resp, as = "parsed", type = "application/json")
  if (length(result) == 0) return(NULL)
  bind_rows(lapply(result, function(x) {
    data.frame(focal_snp   = rsid,
               partner_snp = if (!is.null(x$variation2)) x$variation2 else NA_character_,
               r2          = as.numeric(x$r2),
               d_prime     = as.numeric(x$d_prime),
               population  = population,
               stringsAsFactors = FALSE)
  }))
}
```

```{r fetch_ld, cache=TRUE}
ld_results <- list()

# Validation SNPs — query in primary selection population
for (i in seq_len(nrow(VALIDATION_SNPS))) {
  rs  <- VALIDATION_SNPS$rsid[i]
  pop <- VALIDATION_SNPS$ld_query_pop[i]
  ld  <- fetch_ld(rs, pop, window_kb = 300, r2_min = 0.1)
  if (!is.null(ld)) {
    ld$gene     <- VALIDATION_SNPS$gene[i]
    ld$sel_type <- VALIDATION_SNPS$sel_type[i]
    ld_results[[length(ld_results) + 1]] <- ld
  }
}

# Background SNPs — query in EUR for comparison
for (rs in BACKGROUND_RSIDS) {
  ld <- fetch_ld(rs, "EUR", window_kb = 300, r2_min = 0.1)
  if (!is.null(ld)) {
    ld$gene     <- rs
    ld$sel_type <- "background"
    ld_results[[length(ld_results) + 1]] <- ld
  }
}

ld_all <- bind_rows(ld_results)
cat("Total LD variant pairs retrieved:", nrow(ld_all), "\n")
```

### 7.1 Haplotype extent scores

The **haplotype extent score** is the fraction of SNPs within 300 kb in strong LD (r² ≥ 0.8) with the focal variant. Alleles under recent positive selection are expected to have high scores; background variants and the balancing-selection HBB allele are expected to be lower.

```{r haplotype_scores}
hap_scores <- ld_all %>%
  group_by(gene, focal_snp, sel_type) %>%
  summarise(
    n_partners      = n(),
    frac_r2_hi      = mean(r2 >= 0.8, na.rm = TRUE),
    median_r2       = median(r2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(frac_r2_hi))

kable(hap_scores,
      col.names = c("Gene","rsID","Selection type",
                    "Partner SNPs (300 kb)","Frac r² ≥ 0.8","Median r²"),
      digits = 3,
      caption = "Haplotype extent scores: fraction of partners in strong LD (r² ≥ 0.8) within 300 kb")
```

### 7.2 LD distribution by locus

```{r ld_plot, fig.height=6}
if (nrow(ld_all) > 0) {
  ld_all %>%
    mutate(category = case_when(
      sel_type == "background" ~ "Background",
      sel_type == "balancing"  ~ paste0(gene, "\n(balancing)"),
      TRUE                     ~ paste0(gene, "\n(positive)")
    )) %>%
    ggplot(aes(x = r2, fill = sel_type, color = sel_type)) +
    geom_density(alpha = 0.3, adjust = 1.3) +
    facet_wrap(~ gene, ncol = 4) +
    scale_fill_manual(values  = c(background = "steelblue",
                                   positive   = "#D7191C",
                                   balancing  = "#FF7F00"),
                      name = "Selection type") +
    scale_color_manual(values = c(background = "steelblue",
                                   positive   = "#D7191C",
                                   balancing  = "#FF7F00"),
                       name = "Selection type") +
    labs(title = "r² distribution within 300 kb of each focal SNP",
         subtitle = "Right-skewed (high r²) = extended haplotype = evidence of recent selective sweep",
         x = "r² with focal SNP", y = "Density") +
    theme_classic(base_size = 9) +
    theme(strip.text = element_text(face = "bold", size = 7),
          legend.position = "bottom")
}
```

**Interpretation:**

- **Positive selection loci (LCT, DARC, EDAR, etc.):** Expected to show a right-skewed r² distribution — many nearby SNPs are in strong LD because the sweep was recent and recombination has not yet broken up the haplotype block.
- **HBB (balancing selection):** Expected lower haplotype extent score. The allele has been maintained at intermediate frequency for thousands of generations, allowing recombination to erode LD on both the sickle-cell and normal haplotypes.
- **Background SNPs:** Expected to show a broad or left-skewed r² distribution, reflecting normal background LD levels without a sweep signal.

---

```{r session_info, collapse=TRUE}
sessionInfo()
```
