---
title: "Lab 5"
output: html_document
author: Cole M. in conjunction with Claude Code
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

---

## Question 1: Cancer Risk vs. Cell Divisions

### 1A: Download Data and Plot (log10-log10 scale)

```{r q1a}
cancerFile <- "cancerRisk.txt"
if (!file.exists(cancerFile)) {
  download.file("https://fodorclasses.github.io/classes/stats2020/cancerRisk.txt",
                cancerFile)
}

cancer <- read.table(cancerFile, header = TRUE, sep = "\t")

plot(log10(cancer$CumulativeCellDivisions), log10(cancer$Lifetime_cancer_risk),
     xlab = "log10(Cumulative Cell Divisions)",
     ylab = "log10(Lifetime Cancer Risk)",
     main = "Cancer Risk vs. Cumulative Cell Divisions (log10-log10)",
     pch  = 16, col = "steelblue")
```

This reproduces Figure 1 from Tomasetti & Vogelstein (2015). Each point is a
cancer type; the axes are both on a log10 scale.

---

### 1B: Linear Model and Regression Line

```{r q1b}
myLm <- lm(log10(Lifetime_cancer_risk) ~ log10(CumulativeCellDivisions),
           data = cancer)

plot(log10(cancer$CumulativeCellDivisions), log10(cancer$Lifetime_cancer_risk),
     xlab = "log10(Cumulative Cell Divisions)",
     ylab = "log10(Lifetime Cancer Risk)",
     main = "Cancer Risk vs. Cumulative Cell Divisions with Regression Line",
     pch  = 16, col = "steelblue")
abline(myLm, col = "red", lwd = 2)
legend("topleft", legend = "Regression line", col = "red", lwd = 2)
```

---

### 1C: P-value and R-squared

```{r q1c}
lm_summary <- summary(myLm)

pval   <- anova(myLm)$"Pr(>F)"[1]
rsq    <- lm_summary$r.squared

cat("P-value for slope = 0:", pval, "\n")
cat("R-squared:", rsq, "\n")
```

The p-value for the null hypothesis that the slope is zero is extremely small
(far below 0.05), providing very strong evidence that cumulative cell divisions
are associated with lifetime cancer risk. The R-squared value indicates that
roughly **`r round(rsq * 100, 1)`%** of the variance in log10 cancer risk is
explained by log10 cumulative cell divisions.

---

### 1D: Residual Diagnostics

```{r q1d}
par(mfrow = c(1, 2))

# Residuals vs Fitted (constant variance check)
plot(myLm$fitted.values, myLm$residuals,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs. Fitted",
     pch  = 16, col = "steelblue")
abline(h = 0, col = "red", lty = 2)

# Q-Q plot (normality check)
qqnorm(myLm$residuals, pch = 16, col = "steelblue",
       main = "Normal Q-Q Plot of Residuals")
qqline(myLm$residuals, col = "red", lwd = 2)

par(mfrow = c(1, 1))
```

**Constant variance:** The residuals vs. fitted plot shows no strong funnel
shape or systematic pattern, suggesting the constant-variance assumption is
roughly satisfied on the log10-log10 scale.

**Normality:** The Q-Q plot follows the reference line reasonably well in the
center, though there is some deviation in the tails. With only
`r nrow(cancer)` data points (cancer types), slight deviations are common and
the normality assumption appears broadly reasonable. Overall, the model
assumptions are defensible for these data.

---

## Question 2: Microbiome BMI Association

```{r q2_load}
ccFile  <- "caseControlData.txt"
bmiFile <- "BMI_Data.txt"

if (!file.exists(ccFile)) {
  download.file("http://afodor.github.io/classes/stats2015/caseControlData.txt",
                ccFile)
}
if (!file.exists(bmiFile)) {
  download.file("http://afodor.github.io/classes/stats2015/BMI_Data.txt",
                bmiFile)
}

ccData  <- read.table(ccFile,  header = TRUE, sep = "\t", comment.char = "",
                      check.names = FALSE)
bmiData <- read.table(bmiFile, header = TRUE, sep = "\t", comment.char = "")
```

```{r q2_merge}
# Process sample IDs in ccData to match BMI file keys
sampleIDs <- as.character(ccData[, 1])

makeKey <- function(sampleID) {
  key <- sub("case",    "", sampleID)
  key <- sub("control", "", key)
  key <- strsplit(key, "_")[[1]][1]
  return(key)
}

keys <- sapply(sampleIDs, makeKey)

# Match keys to BMI table
bmiRownames <- as.character(bmiData[, 1])
matchIdx    <- match(keys, bmiRownames)

bmiValues <- bmiData[matchIdx, "bmi"]

# Keep only samples with a BMI value
keep      <- !is.na(bmiValues)
bmiVec    <- bmiValues[keep]
ccSubset  <- ccData[keep, ]

cat("Samples with BMI data:", sum(keep), "\n")
```

```{r q2_regression}
# OTU columns: all columns except the first (sample ID)
otuCols <- 2:ncol(ccSubset)

# Convert OTU counts to relative abundance (row-wise)
otuCounts   <- as.matrix(ccSubset[, otuCols])
rowTotals   <- rowSums(otuCounts)
otuRelAbund <- otuCounts / rowTotals

pvals <- numeric(length(otuCols))
otuNames <- colnames(ccSubset)[otuCols]

for (i in seq_along(otuCols)) {
  relAbund <- otuRelAbund[, i]
  myLm_otu <- lm(relAbund ~ bmiVec)
  pvals[i] <- anova(myLm_otu)$"Pr(>F)"[1]
}

names(pvals) <- otuNames
```

```{r q2_plot}
hist(pvals,
     breaks = 50,
     main   = "P-values: Linear Regression of BMI vs. OTU Relative Abundance",
     xlab   = "p-value",
     col    = "steelblue")
```

```{r q2_fdr}
adjPvals <- p.adjust(pvals, method = "BH")
sigOTUs  <- sum(adjPvals < 0.10, na.rm = TRUE)
cat("OTUs significant at 10% FDR:", sigOTUs, "\n")
cat("Total OTUs tested:", length(pvals), "\n")
```

**Are p-values uniformly distributed?** The histogram of p-values appears
roughly uniform across [0, 1], consistent with the null hypothesis being true
for most OTUs. There is no strong spike near zero that would indicate a large
number of true associations.

**Does the microbial community appear to influence body weight?** Based on
these results, there is little evidence that OTU relative abundances are
strongly associated with BMI in this cohort. The distribution of p-values does
not depart dramatically from uniformity.

**Significant at 10% FDR?** After Benjamini-Hochberg correction,
**`r sigOTUs`** OTU(s) are significant at a 10% false discovery rate,
suggesting minimal or no reliable associations between BMI and the microbial
community in this dataset.
