---
title: "lab_1"
output: pdf_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls())
```

# Use the mean and sd functions to calculate the mean and standard deviation of the following numbers  (2.3,  4.3 ,1.2 , 3.4, 8.3, 12.2) 

``` {r}
nums <- c(2.3,  4.3 ,1.2 , 3.4, 8.3, 12.2)
mean <- mean(nums)
sd <- sd(nums)
```

# Use mean and sd in R to calculate the average and sd of (4,5,6 …. 678,679,680) [the set of every positive integer between 4 and 680]

``` {r}
nums2 <- seq(4:680)
avg <- mean(nums2)
sd <- sd(nums2)
```

# (3) Suppose you are given GC content for sequences from a human genome with values for many sequences: gc <- runif(1000, min = 0.2, max = 0.8). 
- Plot a histogram of gc.
- Compute the mean and standard deviation.
- Does this look like a reasonable sequencing dataset? Why or why not?

```{r}
gc <- runif(1000, min = 0.2, max = 0.8)
hist(gc)
mean(gc)
sd(gc)
```
This distribution is not good as the numbers are basically "random" this shows that you are looking at numbers that have been generated and is not accurate to a real gene set. 
