---
title: "Lab4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Setup: Download and Load Data

```{r load_data}
zipUrl  <- "http://afodor.github.io/classes/stats2015/longitdunalRNASeqData.zip"
zipFile <- "longitdunalRNASeqData.zip"
dataFile <- "nc101_scaff_dataCounts.txt"

if (!file.exists(dataFile)) {
  download.file(zipUrl, zipFile)
  unzip(zipFile)
}

myT <- read.table(dataFile, header = TRUE, row.names = 1)
```

---

## Question 2: log10-log10 Plot of D2_01 vs D2_02

```{r q2}
# Add 1 to avoid log10(0)
plot(log10(myT$D2_01 + 1), log10(myT$D2_02 + 1),
     xlab = "log10(D2_01 + 1)",
     ylab = "log10(D2_02 + 1)",
     main = "Gene Expression: D2_01 vs D2_02 (log10 scale)",
     pch  = 16, cex = 0.5, col = rgb(0, 0, 0, 0.3))
abline(0, 1, col = "red", lwd = 2)
```

**Qualitative assessment:** The two biological replicates show very similar
patterns of gene expression. Points fall tightly along the identity line,
indicating that genes expressed at high levels in D2_01 are also expressed at
high levels in D2_02, as expected for biological replicates from the same
experimental condition.

---

## Question 3: Variance vs Mean (log10-log10 scale)

```{r q3}
geneMeans <- apply(myT, 1, mean)
geneVars  <- apply(myT, 1, var)

# Keep genes with positive mean and variance for log plotting
keep <- geneMeans > 0 & geneVars > 0

plot(log10(geneMeans[keep]), log10(geneVars[keep]),
     xlab = "log10(Mean)",
     ylab = "log10(Variance)",
     main = "Variance vs Mean across all genes (log10 scale)",
     pch  = 16, cex = 0.5, col = rgb(0, 0, 0, 0.3))

# Identity line: variance = mean
abline(0, 1, col = "red", lwd = 2)
legend("topleft", legend = "Identity (var = mean)", col = "red", lwd = 2)
```

**Does the mean equal the variance?** No. The points fall substantially
**above** the red identity line, meaning the variance is much larger than the
mean for most genes. This is characteristic of **overdispersion** — a
hallmark of RNA-seq count data that violates the Poisson assumption
(where mean = variance). A Negative Binomial model would be more appropriate.

---

## Question 4: Fisher Test for First Gene (NC101_00003)

```{r q4}
firstGene <- "NC101_00003"

# Counts assigned to this gene
a <- myT[firstGene, "D2_01"]   # assigned in D2_01
b <- myT[firstGene, "D2_02"]   # assigned in D2_02

# Total sequences in each sample
total_D2_01 <- sum(myT$D2_01)
total_D2_02 <- sum(myT$D2_02)

# 2x2 contingency table
#                 D2_01                D2_02
# Assigned        a                    b
# Not assigned    total_D2_01 - a      total_D2_02 - b

contTable <- matrix(
  c(a,               b,
    total_D2_01 - a, total_D2_02 - b),
  nrow = 2, byrow = TRUE,
  dimnames = list(
    c("Assigned to NC101_00003", "Not assigned"),
    c("D2_01", "D2_02")
  )
)

print(contTable)

fisher_result <- fisher.test(contTable, alternative = "two.sided")
cat("Fisher test p-value for NC101_00003:", fisher_result$p.value, "\n")
```

---

## Question 5: Fisher Test for All Genes

```{r q5_all_genes}
total_D2_01 <- sum(myT$D2_01)
total_D2_02 <- sum(myT$D2_02)

fisher_pvals <- apply(myT, 1, function(row) {
  a <- row["D2_01"]
  b <- row["D2_02"]
  mat <- matrix(c(a, b, total_D2_01 - a, total_D2_02 - b),
                nrow = 2, byrow = TRUE)
  fisher.test(mat, alternative = "two.sided")$p.value
})

hist(fisher_pvals,
     breaks = 50,
     main   = "Histogram of Fisher Test P-values (all genes)",
     xlab   = "p-value",
     col    = "steelblue")
```

**Are p-values uniformly distributed?** No. There is a strong spike near
p = 1, with very few small p-values. This is the **opposite** of what we
would expect if all null hypotheses were true (uniform distribution).
The p-values are less significant than expected — most genes show no
evidence of differential expression between the two biological replicates,
consistent with them being from the same condition.

```{r q5_filtered}
myT_filt <- myT[(myT$D2_01 + myT$D2_02 > 50), ]
cat("Genes remaining after abundance filter:", nrow(myT_filt), "\n")

total_D2_01_f <- sum(myT_filt$D2_01)
total_D2_02_f <- sum(myT_filt$D2_02)

fisher_pvals_filt <- apply(myT_filt, 1, function(row) {
  a <- row["D2_01"]
  b <- row["D2_02"]
  mat <- matrix(c(a, b, total_D2_01_f - a, total_D2_02_f - b),
                nrow = 2, byrow = TRUE)
  fisher.test(mat, alternative = "two.sided")$p.value
})

hist(fisher_pvals_filt,
     breaks = 50,
     main   = "Fisher Test P-values (genes with count > 50)",
     xlab   = "p-value",
     col    = "coral")
```

**Effect of removing low-abundance genes:** After filtering, the distribution
becomes more uniform-ish, though still somewhat right-skewed toward p = 1.
Low-abundance genes have very low counts in both samples, so nearly any small
difference looks proportionally large, inflating p-values toward 1 or
producing very significant p-values when one sample has 0 reads.
Removing them makes the test more meaningful.

---

## Question 6: Poisson Test for First Gene (NC101_00003) with Pseudo-count

```{r q6}
myT_pseudo <- myT + 1

firstGene <- "NC101_00003"

total_D2_01_ps <- sum(myT_pseudo$D2_01)
total_D2_02_ps <- sum(myT_pseudo$D2_02)

count_D2_01 <- myT_pseudo[firstGene, "D2_01"]
count_D2_02 <- myT_pseudo[firstGene, "D2_02"]

# Expected frequency (rate) estimated from D2_01
p_hat <- count_D2_01 / total_D2_01_ps

# Expected count in D2_02 under null: p_hat * total_D2_02
lambda_expected <- p_hat * total_D2_02_ps

cat("Counts in D2_01:", count_D2_01, "\n")
cat("Total D2_01:", total_D2_01_ps, "\n")
cat("Expected frequency p =", p_hat, "\n")
cat("Counts in D2_02:", count_D2_02, "\n")
cat("Expected count in D2_02:", lambda_expected, "\n")

poisson_result <- poisson.test(count_D2_02, T = total_D2_02_ps / total_D2_01_ps,
                                r = count_D2_01, alternative = "two.sided")
# Simpler: test if D2_02 count matches expected rate
poisson_result2 <- poisson.test(count_D2_02, lambda_expected)

cat("Poisson test p-value for NC101_00003:", poisson_result2$p.value, "\n")
```

---

## Question 7: Poisson P-values for All Genes vs Fisher P-values

```{r q7}
total_D2_01_ps <- sum(myT_pseudo$D2_01)
total_D2_02_ps <- sum(myT_pseudo$D2_02)

poisson_pvals <- apply(myT_pseudo, 1, function(row) {
  count1 <- row["D2_01"]
  count2 <- row["D2_02"]
  p_hat  <- count1 / total_D2_01_ps
  lambda <- p_hat * total_D2_02_ps
  poisson.test(count2, lambda)$p.value
})

# Fisher p-values computed on original (no pseudo-count) data
fisher_pvals_all <- apply(myT, 1, function(row) {
  a <- row["D2_01"]
  b <- row["D2_02"]
  mat <- matrix(c(a, b, total_D2_01 - a, total_D2_02 - b),
                nrow = 2, byrow = TRUE)
  fisher.test(mat, alternative = "two.sided")$p.value
})

# Align by gene names
common_genes <- intersect(names(poisson_pvals), names(fisher_pvals_all))
fp <- fisher_pvals_all[common_genes]
pp <- poisson_pvals[common_genes]

plot(log10(fp + 1e-300), log10(pp + 1e-300),
     xlab = "log10(Fisher p-value)",
     ylab = "log10(Poisson p-value)",
     main = "Fisher vs Poisson P-values (log10 scale)",
     pch  = 16, cex = 0.5, col = rgb(0, 0, 0, 0.3))
abline(0, 1, col = "red", lwd = 2)
legend("topleft", legend = "Identity", col = "red", lwd = 2)

cor_val <- cor(log10(fp + 1e-300), log10(pp + 1e-300), method = "spearman")
cat("Spearman correlation between log10 p-values:", round(cor_val, 4), "\n")
```

**How well do they agree?** The two sets of p-values are moderately correlated
along the identity line, but the agreement is imperfect. Genes flagged as
significant by one test are often also flagged by the other, but there is
substantial scatter. This is expected because both models make assumptions
unsupported by the data:

- **Fisher test** assumes a fixed total library size and tests for proportional
  differences, but does not account for overdispersion.
- **Poisson test** assumes mean = variance, which we showed in Question 3 is
  violated (variance >> mean due to overdispersion in RNA-seq data).

With n = 1 replicate per condition and overdispersed counts, both models are
misspecified and conclusions should be interpreted with caution.
