---
title: "Lab 6"
output: html_document
author: Cole M. in conjunction with Claude Code
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Setup: Download, Load, Filter, and Normalize Data

```{r load_data}
zipUrl   <- "http://afodor.github.io/classes/stats2015/longitdunalRNASeqData.zip"
zipFile  <- "longitdunalRNASeqData.zip"
dataFile <- "nc101_scaff_dataCounts.txt"

if (!file.exists(dataFile)) {
  download.file(zipUrl, zipFile)
  unzip(zipFile)
}

myT <- read.table(dataFile, header = TRUE, row.names = 1, sep = "\t")

# Remove rare genes (row median <= 5)
myT <- myT[apply(myT, 1, median) > 5, ]

# Column-sum normalize each sample to relative abundance
myTNorm <- myT
for (i in 1:ncol(myT)) {
  colSum <- sum(myT[, i])
  myTNorm[, i] <- myTNorm[, i] / colSum
}

# Category and time vectors
# Columns: 1-3 = day 2, 4-6 = week 12, 7-11 = week 18 (labeled w20)
cats <- factor(c(rep("day2", 3), rep("week12", 3), rep("week18", 5)),
               levels = c("day2", "week12", "week18"))
times <- c(rep(2, 3), rep(86, 3), rep(128, 5))

cat("Genes retained after filtering:", nrow(myTNorm), "\n")
cat("Samples:", ncol(myTNorm), "\n")
```

---

## Question A: One-way ANOVA (three categories)

For each gene, fit a one-way ANOVA with factor `cats` (day2 / week12 / week18).

```{r qA}
pValuesOneWayAnova <- vector(length = nrow(myTNorm))
index              <- vector(length = nrow(myTNorm))

for (i in 1:nrow(myTNorm)) {
  index[i] <- i
  myData   <- as.numeric(myTNorm[i, ])
  myLm     <- lm(myData ~ cats)
  pValuesOneWayAnova[i] <- anova(myLm)$"Pr(>F)"[1]
}

hist(pValuesOneWayAnova,
     breaks = 50,
     main   = "One-way ANOVA p-values (three categories)",
     xlab   = "p-value",
     col    = "steelblue")

adjA   <- p.adjust(pValuesOneWayAnova, method = "BH")
sigA   <- sum(adjA < 0.05, na.rm = TRUE)
cat("Genes significant at BH FDR < 0.05 (Model A):", sigA, "\n")
```

There is a strong spike of small p-values, indicating many genes differ across
the three time categories. **`r sigA`** genes are significant at a
BH-corrected 5% FDR.

---

## Question B: Linear Regression on Time (days)

For each gene, fit a linear regression with time in days (2, 86, 128).

```{r qB}
pValuesRegression <- vector(length = nrow(myTNorm))

for (i in 1:nrow(myTNorm)) {
  myData <- as.numeric(myTNorm[i, ])
  myLm   <- lm(myData ~ times)
  pValuesRegression[i] <- anova(myLm)$"Pr(>F)"[1]
}

hist(pValuesRegression,
     breaks = 50,
     main   = "Regression on time p-values (two-parameter model)",
     xlab   = "p-value",
     col    = "darkorange")

adjB <- p.adjust(pValuesRegression, method = "BH")
sigB <- sum(adjB < 0.05, na.rm = TRUE)
cat("Genes significant at BH FDR < 0.05 (Model B):", sigB, "\n")
```

**`r sigB`** genes are significant at a BH-corrected 5% FDR under the
two-parameter linear regression on time.

---

## Question C: Comparing the Three-Parameter and Two-Parameter Models

For each gene, test whether the three-parameter categorical model (A) provides
a significantly better fit than the two-parameter regression-on-time model (B).
Because (B) is nested in (A) (the categorical model has one extra parameter),
we can compute a nested F-statistic and obtain the p-value with `pf`.

```{r qC}
pValueModelDiff <- vector(length = nrow(myTNorm))

n <- ncol(myTNorm)   # 11 samples
# df for full (categorical) model:   n - 3
# df for reduced (regression) model: n - 2
dfFull <- n - 3
dfDiff <- 1          # (n-2) - (n-3) = 1

for (i in 1:nrow(myTNorm)) {
  myData  <- as.numeric(myTNorm[i, ])
  lmFull  <- lm(myData ~ cats)     # three-parameter
  lmRed   <- lm(myData ~ times)    # two-parameter

  rssFull <- sum(residuals(lmFull)^2)
  rssRed  <- sum(residuals(lmRed)^2)

  fstat   <- ((rssRed - rssFull) / dfDiff) / (rssFull / dfFull)
  # One-sided upper tail of F-distribution
  pValueModelDiff[i] <- pf(fstat, dfDiff, dfFull, lower.tail = FALSE)
}

hist(pValueModelDiff,
     breaks = 50,
     main   = "Model comparison p-values (A vs. B)",
     xlab   = "p-value",
     col    = "seagreen")

adjC <- p.adjust(pValueModelDiff, method = "BH")
sigC <- sum(adjC < 0.05, na.rm = TRUE)
cat("Genes where Model A significantly improves on Model B (BH FDR < 0.05):",
    sigC, "\n")
```

**`r sigC`** genes show a statistically significant improvement of the
three-parameter categorical model over the two-parameter regression-on-time
model at a 5% FDR.

---

## Question D: Top-Hit Plots Under Each Model

Assemble a data frame of p-values and row indices, then plot the most
significant gene under each of the three models.

```{r qD_frame}
myFrame <- data.frame(index,
                      pValuesOneWayAnova,
                      pValuesRegression,
                      pValueModelDiff)
```

### (A) Top hit — one-way ANOVA (categorical)

```{r qD_A}
frameA  <- myFrame[order(myFrame$pValuesOneWayAnova), ]
topA    <- frameA$index[1]
topAgene <- rownames(myTNorm)[topA]

boxplot(as.numeric(myTNorm[topA, ]) ~ cats,
        xlab = "Time category",
        ylab = "Relative abundance",
        main = paste0("Top one-way ANOVA hit: ", topAgene,
                      "\np = ", signif(pValuesOneWayAnova[topA], 3)),
        col  = c("#A6CEE3", "#1F78B4", "#08306B"))
stripchart(as.numeric(myTNorm[topA, ]) ~ cats,
           vertical = TRUE, add = TRUE, pch = 16, col = "red", method = "jitter")
```

### (B) Top hit — regression on time

```{r qD_B}
frameB   <- myFrame[order(myFrame$pValuesRegression), ]
topB     <- frameB$index[1]
topBgene <- rownames(myTNorm)[topB]

yB  <- as.numeric(myTNorm[topB, ])
lmB <- lm(yB ~ times)

plot(times, yB,
     xlab = "Time (days)",
     ylab = "Relative abundance",
     main = paste0("Top regression hit: ", topBgene,
                   "\np = ", signif(pValuesRegression[topB], 3)),
     pch  = 16, col = "darkorange")
abline(lmB, col = "red", lwd = 2)
legend("topright", legend = "Regression line", col = "red", lwd = 2, bty = "n")
```

### (C) Top hit — model comparison (A vs. B)

```{r qD_C}
frameC   <- myFrame[order(myFrame$pValueModelDiff), ]
topC     <- frameC$index[1]
topCgene <- rownames(myTNorm)[topC]

boxplot(as.numeric(myTNorm[topC, ]) ~ cats,
        xlab = "Time category",
        ylab = "Relative abundance",
        main = paste0("Top model-comparison hit: ", topCgene,
                      "\np = ", signif(pValueModelDiff[topC], 3)),
        col  = c("#B2DF8A", "#33A02C", "#1B5E20"))
stripchart(as.numeric(myTNorm[topC, ]) ~ cats,
           vertical = TRUE, add = TRUE, pch = 16, col = "red", method = "jitter")
```

---

## Question E: Which Model is More Appropriate?

The two-parameter linear regression on time (Model B) assumes that relative
abundance changes **linearly** with day. The three-parameter categorical model
(Model A) makes no such assumption — each time point gets its own mean.

Looking at the results:

- Model A flags **`r sigA`** genes at 5% FDR; Model B flags **`r sigB`**.
- The model-comparison test in (C) finds **`r sigC`** genes where Model A
  provides a significantly better fit than Model B.

Because the sampling is very unevenly spaced in time (day 2, day 86, day 128),
a true linear response in days is biologically implausible for most genes —
abundances typically plateau, rebound, or shift non-monotonically. The top-hit
boxplots under Model A and Model C show patterns (e.g. week 12 markedly
different from both day 2 and week 18) that a straight line through day cannot
capture. The large number of genes where Model A significantly outperforms
Model B in (C) confirms this.

**The three-parameter categorical model (A) is the more appropriate model for
these data**, because it does not impose a linear-in-time assumption that the
biology does not support.
